---
title: "Monetary policy and asset pricing - Assignment no.1"
author: "Riccardo Dal Cero"
date: "06/11/2022"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial Operations

```{r p0, message=FALSE}
# Clear the variables
rm(list = ls())

# Install packages
packages <- c("matlib", "rmarkdown", "tinytex")
new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages)
invisible(lapply(packages, library, character.only = TRUE))
# Load packages
library(matlib)
options(digits = 15)

```

## POINT 2

We define a bisection alghorithm that max the value function which represents
the following preferences: risk premium rate > 4% and risk free rate -> Epsilon

### Risk computation
the risk computation function take m as input and compute the risk free / premium rate and recall
the value function with such values.
```{r p1, include=TRUE}
riskComputation <- function(m,param){
    #parameters 
    phi     <-  param[4]
    beta    <- 0.96
    gamma   <- param[3]
    h       <- param[1]
    l       <- param[2]
    epsilon <- 0.012

    #Generate pi Matrix
    valuesPi        <-   c(phi, epsilon, 1 - phi - epsilon, 1 / 2,
     0, 1 / 2, 1 - phi - epsilon, epsilon, phi)
    Pi              <-   matrix(valuesPi, byrow = TRUE, nrow = 3, ncol = 3)


    #a vector that contains the three states of the system
    values_states   <-   c(h,m,l)

    #ouput vector
    out <-  c()

    #stationary prob.
    st_prob <- c(0.5, epsilon, 0.5) * (1 + epsilon)

    #Computing the zero-cupon bond 
    q   <-  beta * (Pi %*% (values_states ^- gamma))

    #compute risk-free rate vector
    risk_free   <-   (1 / (q)) - 1

    #compute mean risk free rate using stationary prob. as a weight
    rf <- risk_free[, 1] %*% st_prob
    as.numeric(rf)

    #risky_rate
    values_matrix <- diag(3) * (values_states^(1 - gamma))

    #Computing Pi star
    Pi_star <-  Pi %*% values_matrix

    I       <- diag(3) #diagonal matrix
    v1      <- c(1, 1, 1) #[1,1,1]

    #Computing risk premium prices 
    p       <- (inv(I - beta * Pi_star)) %*% (beta * (Pi_star %*% v1))
    
    #Converting prices in rates

    # values in diagonal matrix
    values     <- c((1 + p) * values_states)

    #generate the diagonal matrix
    v2      <- matrix(diag(values), 3, 3)

    #generate the matrix
    v3      <- matrix((p)^(-1), 3, 3)
    
    #computing the matrix to convert the prices into rates
    v4      <- v3 %*% v2

    #computing the avarage risk rate using an equal weight for the three status
    rr      <- t(c(rep(1 / 3, 3))) %*% v4 %*% st_prob
    as.numeric(rr)

    #compute index
    i       <- v((rr - 1) * 100, rf * 100)

    out     <- c(rf * 100, (rr - 1) * 100, log(i))

    return(out)

}


```



### Value function
Value function: given the risk premium rate = 'dummy' and 'x' = risk 
free rate as input return a value which is higher when x -> 1 and
dummy > 4
```{r p3}

v <- function(dummy, x) {

  for (t in seq_along(dummy)) {
    if (dummy[t] > 4) {
      dummy[t]  <-  2
    }
    else {
      dummy[t]  <-  1
    }
  }
  value <-  abs(1 / (1 - x)) * dummy
  return(value)
}
```

### Bisection alghorithm

```{r p4, dev='svg'}
bisection <- function(param) {
    #bisection variable
    #setting the max min value for m
    max     <- 20.0000
    min     <- 0.0000
    half    <- (max - min) / 2 + min
    i       <- 0 #values from value function
    count   <- 1 #numeber of iteration
    out <-c()
    while (i < 20) {
      print(cbind("Iteration: ", count))

      #A vector which takes the first quartile and the last
      #from the set of possibile values (max-min)
      m1  <- c((max - min) / 4, (max - min) * 3 / 4) + min

      bis <- matrix(NA, nrow = 2, ncol = 4)
      #   index    m     free    p   val
      #   1       m1[1]
      #   2       m1[2]

      bis[, 1] <- as.numeric(m1)

      #recall a function to compute risk rate for m[1] and the index
      bis[1, 2:4] <- riskComputation(m1[1], param)

      #recall a function to compute risk rate for m[1] and the index
      bis[2, 2:4] <- riskComputation(m1[2], param)
      
      #select the one with max index(value)
      if (bis[1, 4] > bis[2, 4]) {
        #the m in the first quartile is the one with greater index(value)
        #so the m value that max the value function is in the first half
        #of the distribution
        #so now I will restrict the possibile values of m by half, selecting
        #the first half of the distribution

        max     <- half
        half    <- (max - min) / 2 + min
        print(bis[1,])
        i       <- bis[1, 4]  #output of the value function given m
        out <- bis[1,]

      } else if (bis[1, 4] < bis[2, 4]) {
        #the same as above, but we take the other half of the distrib.

        min     <- half
        half    <- (max - min) / 2 + min
        print(bis[2,])
        i       <- bis[2, 4]
        out <-  bis[2,]

      } else {
        #In case the value function return the same value for the two m
        #we have convergence so I stop the cycle

        print(bis[1, ])

        i <- 100
      }
      count <- count + 1
    }
    return(out)
}
```

### Main function

```{r p5, echo=TRUE}
main <- function() {
  phi     <- 0.43
  gamma   <- 2
  h       <- 1.054
  l       <- 0.982
  param   <- c(h, l, gamma, phi)
  m       <- bisection(param)
  return(m)
}
main()
```