---
title: "Monetary Policy and Asset Pricing - Assignment no.3"
author: "Federico Vicentini"
date: "06/11/2022"
output: html_document
bibliography: references.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r p0, message=FALSE, echo=FALSE}
# Clear the variables
rm(list = ls())

# Install packages
packages <- c("matlib", "rmarkdown", "tinytex","plotly","rgl","knitr")
new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages)
invisible(lapply(packages, library, character.only = TRUE))
# Load packages
library(matlib)
library(scatterplot3d)
library(plotly)
library(rgl)
library(knitr)
options(digits = 15) 

```
## POINT 1
```{r p100,echo=TRUE}
data <- matrix(NA, nrow = 5, ncol = 10)

a <- c(0.023, 0.041, 0.2)
b <- c(0.015, 0.030, .9)
c <- c(0.030, .05, -.2)
d <- c(.018, .036, .0)
e <- c(.025, .045, -0.9)


data[1,1:3 ] <- a
data[2,1:3 ] <- b
data[3,1:3 ] <- c
data[4,1:3 ] <- d
data[5,1:3 ] <- e
col_names <- c("u", "n", "p", "l", "h", "phi", "q_h", "q_l","r_h","r_l")
countries <- c("A", "B", "C", "D", "E")

colnames(data) <- col_names
rownames(data) <- countries

beta <- 0.96
gam <- 2
#compute the high and low values 
for (count in 1:5) {
   u <- data[count,1]
   n <- data[count,2]
   p <- data[count,3]
   data[count, 4] <- l <- 1 + u - n
   data[count, 5] <- h <- 1 + u + n
   k <- (p + (u / n)^2) * n^2
   data[count, 6] <- k
   data[count, 6] <- phi <- (k - ((h - 1) * (l - 1))) / (0.5 * (h - 1)^2 + 0.5 *
   (l - 1)^2 - ((h - 1) * (l - 1)))
   #risk free rate 
   data[count, 7] = beta * (phi * h^(-gam)+ (1 - phi) * l^(-gam)) #q_h
   data[count, 8] = beta * ((1 - phi) * h^(-gam) + phi * l^(-gam)) #q_l
   data[count, 9] = data[count, 7]^(-1)-1
   data[count, 10] = data[count, 8]^(-1)-1 #q_l
}
kable(data)


```

The code reported computes the transition matrix and the risk-free rates for each country. Comments are reported in the code.

## Comment on the table results:

COUNTRY A:
In country A, since phi=0.6, if an individual finds itself in the state h, he has a higher probability of staying in h,
thus decreasing the demand for risk-free assets and increasing their rate of return. The reverse happens if you find yourself
in state l, as demand for risk-free assets increases and their rate decreases. This is why we see r_h>r_l.

COUNTRY B:
The situation is analogous to the one seen in country A, but phi is higher (0.95), thus leading to higher stability
in economic performance, and a wider difference between r_l and r_h.

COUNTRY C:
In country C, since phi=0.4, if an individual finds itself in the state h, he has a higher probability of ending up in l,
thus increasing the demand for risk-free assets and decreasing their rate of return. The reverse happens if you find yourself
in state l, as demand for risk-free assets decreases and their rate increases. This is why we see r_l>r_h




COUNTRY D:
In country D, since phi=0.5, if an individual finds itself in the state h, he has the same probability of going to l or staying in h,
thus the demand for risk-free assets and risky assets stay the same. If you find yourself in state l, the same reasoning applies.
Thus, as we can clearly see by the results, r_h=r_l


COUNTRY E:
In country E, since phi=0.05, if an individual finds itself in the state h, he has a almost certain probability of ending up in l,
thus increasing the demand for risk-free assets and decreasing their rate of return. The reverse happens if you find yourself
in state l, as demand for risk-free assets decreases and their rate increases. This is why we see r_l>r_h
Also, the gap between the two rates is very wide, since volatility is considerably higher compared to country C. 





## POINT 2

The exercise demands to find a value of the third state of the world (m) which fits the following 2 criterion: a risk-free rate
around 1\% and an equity premium above 4\%.

In order to find the correct value of m without running a brute-force algorithm with thousands of iterations, we 
have defined a bisection alghorithm that maximizes the value function which represents
the exercise's criterion.

### Risk computation
The risk computation function takes m as input, computes the risk-free rate and the risky rate and computes the value
function with those values.
```{r p1, include=TRUE}
riskComputation <- function(m, param){
    #the risk computation function take m as input and
    #compute the risk free / premium rate and recall
    #the value function with such values.
    
    #parameters 
    
    phi     <-   param[4]
    beta    <- 0.96
    gamma   <- param[3]
    h       <- param[1]
    l       <- param[2]
    epsilon <- 0.012

    

    #Generate pi Matrix
    valuesPi        <-   c(phi, epsilon, 1 - phi - epsilon, 1 / 2,
     0, 1 / 2, 1 - phi - epsilon, epsilon, phi)
    Pi              <-   matrix(valuesPi, byrow = TRUE, nrow = 3, ncol = 3)
    Pi

    #a vector that contains the three states of the system
    values_states   <-   c(h,m,l)

    #ouput vector
    out <-  c()

    #stationary prob.
    st_prob <- c(0.5, epsilon, 0.5) * (1 + epsilon)^(-1)
    st_prob

    #Computing the zero-cupon bond 
    q   <-  beta * (Pi %*% (values_states ^- gamma))
    q

    #compute risk-free rate vector
    risk_free   <-   (1 / (q)) - 1

    #compute mean risk free rate using stationary prob. as a weight
    rf <- risk_free[, 1] %*% st_prob
    as.numeric(rf)

    #risky_rate
    values_matrix <- diag(3) * (values_states^(1 - gamma))

    #Computing Pi star
    Pi_star <-  Pi %*% values_matrix
    Pi_star

    I       <- diag(3) #diagonal matrix
    v1      <- c(1, 1, 1) #[1,1,1]

    #Computing risk premium prices 
    p       <- (inv(I - beta * Pi_star)) %*% (beta * (Pi_star %*% v1))
    p
    
    #Converting prices in rates

    # values in diagonal matrix
    values     <- c((1 + p) * values_states)

    #generate the diagonal matrix
    v2      <- matrix(diag(values), 3, 3)

    #generate the matrix
    v3      <- matrix((p)^(-1), 3, 3)
    
    #computing the matrix to convert the prices into rates
    v4      <- v3 %*% v2

    #computing the average risk rate using an equal weight for the three status
    once=matrix(1,3,3)
    once
    v4bis=v4-once
    v4bis
    v5=v4bis*Pi
    v5 #NEW risky rates weighted for conditional
    v6=v5%*%st_prob 
    v6
    rr      <- sum(v6)
    as.numeric(rr)
    
    #rp      <- p[, 1] %*% st_prob
    

    #compute index
    i       <- v(rr * 100, rf * 100)

    out     <- c(rf * 100, rr * 100, i)

    return(out)

}


```



### The Value function
The value function represents our preferences given the risk premium rate = 'dummy' and 'x' = risk-free rate. It returns a higher
value when x -> 1 and dummy > 5. The symmetric hyperbolic  gives the function that evaluates the proximity to 1 of the 
risk-free rate  function:
$$
y=\left|\frac{1}{1-x}\right|
$$
While a dummy variable represents the risk-premium rate condition:
$$
f(n) = \begin{cases} 1, & \mbox{if } risk premium\mbox{ <5\%} \\ 2, & \mbox{otherwise }  \end{cases}
$$
Finally, the value function can be represented as:
$$
g(n) = ln\left(\left|\frac{1}{1-x}\right|*f(n)\right)
$$
```{r p2, include=TRUE}
v <- function(dummy, x) {
  # value function: given the risk premium rate = 'dummy' and
  # 'x' = risk free rate as input return a value which is higher
  # when x -> 1 and  dummy > 4

  for (t in seq_along(dummy)) {
    if (dummy[t] > 5) {
      dummy[t]  <-  2
    }
    else {
      dummy[t]  <-  1
    }
  }
  value <-  log(abs(1 / (1 - x)) * dummy)
  return(value)
}
```

```{r p10, echo=FALSE,message=TRUE}
graphValueFunction  <-  function(){
  risk_prem <-  c(seq(2, 6, 0.25))
  risk_free <-  c(seq(0.01, 2, 0.05))
  value     <- matrix(0, nrow = 0, ncol = 3)
  for (x in risk_prem) {
    for(y in risk_free){
      value  <-  rbind(value, c(v(x, y), x, y ))
    }
     
  }
  value[!is.finite(value[, 1]), 1] <- 0
  fig <- plot_ly( x = ~value[,3], y = ~value[,2], z = ~value[,1],
                marker = list(color = ~value[ ,1], colorscale = c('#FFE1A1', '#ce07b0'), showscale =FALSE))
  fig <- fig %>% add_markers()
  fig <- fig %>% layout(scene = list(xaxis = list(title = 'Risk free rate'),
                                    yaxis = list(title = 'Equity premium rate'),
                                    zaxis = list(title = 'Value')))
fig
}
graphValueFunction()
```

### Bisection alghorithm

The bisection algorithm works within a set max-min range as follows:
1. Cut the range in half
2. Compute the value function in the points located at the limits of the first and fourth quartiles
3. Compare the two values, and select the larger one
4. Restart at 1, with new range set as the selected half of the previous one

The process iterates multiple times, until the value function returns an index greater than 20 (in log terms it is quite high).


```{r p4, dev='svg', echo=TRUE, include=TRUE}
bisection <- function(param) {
    #bisection variable
    #setting the max min value for m
    max     <- 20.0000
    min     <- 0.0000
    half    <- (max - min) / 2 + min
    i       <- 0 #values from value function
    count   <- 1 #numeber of iteration
    out <-c()
    while (i < 20) {
      #print(cbind("Iteration: ", count))

      #A vector which takes the first quartile and the last
      #from the set of possibile values (max-min)
      m1  <- c((max - min) / 4, (max - min) * 3 / 4) + min

      bis <- matrix(NA, nrow = 2, ncol = 4)
      #   index    m     free    p   val
      #   1       m1[1]
      #   2       m1[2]

      bis[, 1] <- as.numeric(m1)

      #recall a function to compute risk rate for m[1] and the index
      bis[1, 2:4] <- riskComputation(m1[1], param)

      #recall a function to compute risk rate for m[1] and the index
      bis[2, 2:4] <- riskComputation(m1[2], param)
      
      #select the one with max index(value)
      if (bis[1, 4] > bis[2, 4]) {
        #the m in the first quartile is the one with greater index(value)
        #so the m value that max the value function is in the first half
        #of the distribution
        #so now I will restrict the possibile values of m by half, selecting
        #the first half of the distribution

        max     <- half
        half    <- (max - min) / 2 + min
        #print(bis[1,])
        i       <- bis[1, 4]  #output of the value function given m
        out <- bis[1,]

      } else if (bis[1, 4] < bis[2, 4]) {
        #the same as above, but we take the other half of the distrib.

        min     <- half
        half    <- (max - min) / 2 + min
        #print(bis[2,])
        i       <- bis[2, 4]
        out <-  bis[2,]

      } else {
        #In case the value function return the same value for the two m
        #we have convergence so I stop the cycle

        #print(bis[1, ])

        i <- 100
      }
      count <- count + 1
    }
    return(out)
}
```


```{r p6, echo=FALSE, include=TRUE, dev='svg'}
graphm <- function(){
  h       <- seq(from = 1.03, to = 1.2, by = 0.01) 
  l       <- seq(from = 0.83, to = 0.99, by = 0.01)

  mlist   <- matrix(0, 1, 3)
  for (i in seq_along(h)){
    for (j in seq_along(l)){
      param <- c(h[i], l[j], gamma, phi)
      v5 <- bisection(param)
      mlist <- rbind(mlist, c(v5[1], h[i], l[j]))
    }
  }

  mlist <-  mlist[mlist[,1]<0.5,]
  mlist <-  mlist[mlist[,1]>0,]
  

  #plot_ly(z = graphmatrix, type = "surface")
  fig <- plot_ly( x = ~mlist[,2], y = ~mlist[,3], z = ~mlist[ ,1],
                marker = list(color = ~mlist[ ,1], colorscale = c('#FFE1A1', '#b311d4'), showscale =FALSE))
  fig <- fig %>% add_markers()
  fig <- fig %>% layout(scene = list(xaxis = list(title = 'h'),
                                    yaxis = list(title = 'l'),
                                    zaxis = list(title = 'm')))
  fig
  #plot(mlist[,2]-mlist[,3], mlist[,1])
  #scatterplot3d(mlist[,2],mlist[,3],mlist[,1])
  
}

```

### Main function

```{r p5, echo=FALSE, include=TRUE}
phi     <- 0.43
gamma   <- 2
h       <- 1.054
l       <- 0.982
param   <- c(h, l, gamma, phi)
m       <- bisection(param)
m
```

Here we perform the main computation of the exercise. The first value displayed is the value of state m, then the values of the
risk-free rate, the risky rate and the value function are reported.
We see at first glance that the value of m is quite low, about 0.399. Initially, we thought that m was a medium state, lying
between h and l. However, we were mistaken, and there is good reason for that.

If m had been lying between h and l, its impact on the equity premium puzzle would have been non-existent, since the variability
of the system would have been already captured by the one between state h and l.
Having a low m serves instead to represent the possibility of having a catastrophic event (like a massive financial crisis), which
erodes your assets by 60\%. This is a good way to eplain the equity premium puzzle. We can see in our results that we were able to 
find a risk-free rate of precisely 1\% and a risky rate of 9.6\% (so, an equity premium of 8.6\%). This fares much better with 
real-world data when compared to Mehra and Prescott's results.

We interpret these improvements as a proof of the fact that catastrophic economic events are taken into consideration by
market participants. In fact, Mehra and Prescott set the value of h and l looking at long-term variability in returns. In this way,
one-period catastrophes are smoothed out, but this is not how economic agents would look at them. In fact, for market participants
the possibility of losing a big chunk of their savings is considered as a real possibility, thus increasing the equity premium
with respect to a model where catastrophic events are not taken into account. 

Note that the equity premium increases more as a result of a lower risk-free rate than of a higher risky rate. This is because a
catastrophic event increases considerably the demand for safer, risk-free assets, thus appreciating them and decreasing the
interest rate. On the other hand, risky assets are purchased only if they present an higher return. This is what drives risky
rates up, but the effect on them is weaker than the one on the risk-free rate.

We made a quick research to check if these results had any corroboration in the economic literature. We were able to find a paper
written by Robert J. Barro [@barro], where the inclusion of possible catastrophic economic occurences is cited as a promising 
path to the resolution of the equity premium puzzle.

Note anyway that even if results are better than those of Mehra and Prescott, the equity premium is still a bit higher than 
what we see in the real US data.



### EXTRA SECTION

```{r p22, echo=FALSE, include=TRUE}
#graphValueFunction()
graphm()
```

We added this final section because we wanted to investigate the relationship between the value of m and the value of h and l.

We interpret h and l as being a measure of the volatility of an economy, and we wanted to see for different values of both
how the value of m found with our alogorithm would change.

We plotted the results using a 3d graph, where we can easily see how an increasing h or an increasing l lead to a lower 
value of m, but the difference between them also plays a role in diminishing m.

You can see this by interacting with the graph: if you consider a point where, for example h=1.15 and l=0.94, we have m=0.35;
in this case we can see how both h and l are higher than in our baseline case, and indeed their difference is higher; thus
we are not surprised to find that m is lower, leading us to believe that a lower value for the catastrphic state is needed
to obtain the desired rates, since the variability of this model is even lower than that of Mehra and Prescott.

Indeed, if you move 1 point in each direction starting from these values, you can find confirmation of this relationship, as
you can see that decreasing h, l or the difference between them will increase the value of m (and vice-versa).


## REFERENCES